---
title: 'Portfolio 10: popdemo'
author: "Justine Lang"
date: "2024-02-15"
output: 
  html_document: 
    code_folding: show
---

```{r setup, echo = TRUE, warning = FALSE, message = FALSE, results = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, results = FALSE)

library(tidyverse)
library(popdemo)
```

Deterministic Projection

```{r}
# look at the tortoise data
Tort
```

```{r}
# run a deterministic projection
tortoise_vector <- runif(8) # generating a random population vector
tortoise_vector <- tortoise_vector/sum(tortoise_vector) #scales the vector to sum to 1
tortoise_projection_100 <- project(Tort, tortoise_vector, time = 100)
print(tortoise_projection_100)
```

```{r}
# extract the separate life-stage counts (just the first 10)
vec(tortoise_projection_100)[1:11, ]

# plot it using the built-in function
plot(tortoise_projection_100)
```


Stochastic Projection

```{r}
# look at the polar bear data
Pbear
```

```{r}
Pbear_vector <- c(0.106, 0.068, 0.106, 0.461, 0.151, 0.108) # from paper
```

```{r}
# run a 50 year stochastic projection with even probability of selecting each matrix
Pbear_projection_random <- project(Pbear, Pbear_vector, time = 50)
plot(Pbear_projection_random)
```

```{r}
# look at the matrices chosen in the above model
Aseq(Pbear_projection_random)
```

```{r}
# specify Aseq matrix with good and bad ice years with their probabilities
p3 <- 0.8 # probability of a bad ice year
bad_years_4x_likely <- matrix(
  rep(c((1-p3)/3, (1-p3)/3, (1-p3)/3, p3/2, p3/2), 5), 5, 5)
bad_years_4x_likely
```

```{r}
# call this matrix into our project() function using Aseq argument
Pbear_projection_badyears4xlikely <- project(Pbear, Pbear_vector,
                                            Aseq = bad_years_4x_likely,
                                            time = 50)
Aseq(Pbear_projection_badyears4xlikely)
plot(Pbear_projection_badyears4xlikely)
```


Task 1:

Define a new matrix to feed into Aseq that takes the previous year’s sea ice status into account.

Make good years more likely if the previous year was good (use 20% likelihood), and conversely # makes bad ice years more likely if the previous year was bad (use 80% likelihood).

```{r}
p4 <- 0.2 # probability next year is a bad year given that this year was a good year
q4 <- 0.8 # probability next year is a bad year given that this year was a bad year
Pbear_task1_matrix <- matrix(c(rep(c((1-p4)/3, (1-p4)/3, (1-p4)/3, p4/2, p4/2), 3),
                      rep(c((1-q4)/3, (1-q4)/3, (1-q4)/3, q4/2, q4/2), 2)), 5, 5)

Pbear_task1_projection <- project(Pbear, Pbear_vector, Aseq = Pbear_task1_matrix, time = 50)
plot(Pbear_task1_projection)
```

Task 2:

Define 2 new parameters which describe the probability next year is a bad year given that this year was a good (use 50% likelihood), and (b) describe the probability next year is a bad year given that this year was a bad year (use 80% likelihood again).

```{r}
p5 <- 0.5 # probability next year is a bad year given that this year was a good year
q5 <- 0.8 # arbitrary object names
Pbear_task2_matrix <- matrix(c(rep(c((1-p5)/3, (1-p5)/3, (1-p5)/3, p5/2, p5/2), 3),
                      rep(c((1-q5)/3, (1-q5)/3, (1-q5)/3, q5/2, q5/2), 2)), 5, 5)

Pbear_task2_projection <- project(Pbear, Pbear_vector, Aseq = Pbear_task2_matrix, time = 50)
plot(Pbear_task1_projection)
```







