---
title: 'Portfolio 4: glarma'
author: "Justine Lang"
date: "2024-02-01"
output: 
  html_document: 
    code_folding: show
---

```{r setup, echo = TRUE, message = FALSE, warning = FALSE, results = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, results = FALSE)

library(glarma)
library(here)
library(dplyr)
library(tidyverse)
```

### Portfolio 4 Questions

Use the ‘glarma’ package in R: Use Bison data. AUer you create your models, visually compare the plots.

a. Which looks like the best fit?

b. Compare the AIC for the models. Which is actually the best fit?

c. Create a model with major events factored in and compare it with the best fit model identified in question 2. Is the model with major events a better fit? Some hints: Think about bison reproduction and how it would affect time lags: When
do they reach sexual maturity? (2 years for females) How long is gestation? (~ 9 months) How often do bison breed? (Once a year)

```{r}
#load in the bison data
bison <- read_csv(here("forcast_glarma", "data", "bison_copy.csv"))
```

```{r}
#add an intercept 
bison$intercept<-as.integer(1)

#Identify variables
y <- bison$bison # y value is bison counts
x0 <- bison %>% select(intercept) 
x0 <- as.matrix(x0) # turns x0 into a matrix 
```

```{r}
#Create three glarma models that have a time lag of 1, 2, and 7
bison_null_lag <- glarma(y, x0, phiLags = c(1), type = "Poi", method = "FS",
                              residuals = "Pearson", maxit = 100, grad = 1e-6)

bison_null_lag2 <- glarma(y, x0, phiLags = c(2), type = "Poi", method = "FS",
                         residuals = "Pearson", maxit = 100, grad = 1e-6)

bison_null_lag7 <- glarma(y, x0, phiLags = c(7), type = "Poi", method = "FS",
                          residuals = "Pearson", maxit = 100, grad = 1e-6)
```


1. Visually compare the plots. Which looks like the best fit?
(Commenting out for knitting purposes because of the error in the one plot)

```{r}
# plot.glarma(bison_null_lag) ###if you have an error in plot.window for the last plot, it's okay to ignore
```

```{r}
# plot.glarma(bison_null_lag2)
```

```{r}
# plot.glarma(bison_null_lag7)
```

Visually, the first one with the time lag of 1 looks like the best fit, though it is a close tie with the second model.

2. Compare the AIC for the models. Which is actually the best fit? 

```{r}
summary(bison_null_lag)
```

```{r}
summary(bison_null_lag2)
```

```{r}
summary(bison_null_lag7)
```

Based on the lowest AIC, the second model (time lag of 2) has the best fit (AIC: 6684.275)

3. Create a model with major events factored in and compare it with the best fit model identified in question 2. Is the model with major events a better fit? 

```{r}
#Add intercept and major events
bison$major_events <- as.integer(0)

bison$major_events[bison$year==1995] <- 1 # reintroduction of wolves to yellowstone NP

bison$major_events[bison$year==1988] <- 1 # 1988 Yellowstone Fire

#Identify variables
x1 <- bison %>% select(intercept, major_events) # x1 explanatory variables, presence of major events
x1 <- as.matrix(x1) #turns x1 into a matrix
```

```{r}
#Create the model with major events. Make the lag the same as the best fit model from question 2 
bison_exp_lag <- glarma(y, x1, phiLags = c(2), type = "Poi", method = "FS", 
                             residuals = "Pearson", maxit = 100, grad = 1e-6)
#summary 
summary(bison_exp_lag)
```

The model with major events factored in has a lower AIC (6265.411), which suggests it's a better fit.

Extra if you're feeling spry 

```{r}
bison_mod <- bison 

bison_mod$est <- bison_exp_lag$fitted.values #adds fitted values

bison_plot<-ggplot(bison_mod, aes(x=year)) + 
  geom_point(col="gray",aes(y=bison),size=3) + 
  geom_point(col="black",aes(y=est),size=3) +
  geom_line(col="black",aes(y=est)) +
  labs(x="Year", y="Bison Population") +
  ylim(0,4100) + xlim(1970,1997)+
  theme_bw()

#Adds two red horizontal lines for the occurrence of fire and reintroduction of wolves
bison_plot <-bison_plot + 
  annotate("rect", xmin = 1988, xmax =1988.1, ymin = 0, ymax = 4100, alpha = .75,fill = "red") + 
  annotate("rect", xmin = 1995, xmax =1995.1, ymin = 0, ymax = 4100, alpha = .75,fill = "red")

bison_plot
```








