---
title: "Portfolio 2: growthcurver"
author: "Justine Lang"
date: "2024-01-16"
output: 
  html_document: 
    code_folding: show
---

```{r setup, echo = TRUE, warning = FALSE, results = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, results = FALSE, message = FALSE)

library(tidyverse)
library(janitor)
library(here)
library(growthcurver) ### older logistic growth package
```

### Portfolio 2 Questions

Use ‘growthcurver’ package in R: Use World data

a. Import the global population data and apply the logistic growth model. Does the model describe the world population well?

b. Choose a country of interest. Does the model describe that country's population well? What countries have logistic growth?

**Exponential Growth of Human Population Example**

```{r}
#get the data
w_pop_data<-read_csv(here("Portfolio1","population-and-demography.csv"))

#clean the data for only the global human population
w_pop_data<- w_pop_data |> clean_names() |>
  filter(country_name=="World") |> 
  select(year, population) |> drop_na()
```

Always plot your data

```{r}
#On the observed scale
human_ts<-ggplot(w_pop_data, aes(x=year, y=population))+
  geom_point()+
  xlab("Year")+
  ylab("Human count (N)")+
  xlim(1949, 2023)+
  ggtitle("Human Population on Earth (1950-2022)")+
  theme_bw()

human_ts
```

```{r}
#on the transformed log scale
human_ts_log<-ggplot(w_pop_data, aes(x=year, y=population))+
  geom_point()+
  xlab("Year")+
  ylab("Human count (N)")+
  scale_y_continuous(trans="log")+ # log scale for linear transformation
  xlim(1949, 2023)+
  ggtitle("Human Population on Earth (1950-2022)")+
  theme_bw()

human_ts_log
```

Use lm() to find estimates

```{r}
human.lm_fit<- lm(log(population)~year,data=w_pop_data)
summary(human.lm_fit) #how well does it fit? 

NO<-exp(human.lm_fit$coefficients[1])
r<-human.lm_fit$coefficients[2]
```

Plot the data and the model

```{r}
# On the transformed scale
human_ts_model_trans<-ggplot(w_pop_data, aes(x=year, y=population))+
  geom_point()+
  xlab("Year")+
  ylab("Human count (N)")+
  scale_y_continuous(trans="log")+
  geom_smooth(method="lm",color="blue")+
  xlim(1949, 2023)+
  ggtitle("Human Population on Earth (1950-2022)")+
  theme_bw()

human_ts_model_trans
```

```{r}
# On the observed scale

# Predicted population values from the model
predicted_df <- data.frame(pop_pred = predict(human.lm_fit, w_pop_data), year=w_pop_data$year)
predicted_df<- predicted_df |> mutate(N_est=exp(pop_pred)) #transformed back to observed population estimates


human_ts_model<-ggplot(w_pop_data, aes(x=year, y=population))+
  geom_point(color="black")+
  geom_line(color='red',data = predicted_df, aes(x=year, y=N_est))+
  xlab("Year")+
  ylab("Human count (N)")+
  xlim(1949, 2023)+
  ggtitle("Human Population on Earth (1950-2022)")+
  theme_bw()

human_ts_model
```

**Exponential Growth of Denmark Population**

```{r}
w_pop <-read_csv(here("Portfolio1","population-and-demography.csv"))

#clean the data for only the global human population
den_data<- w_pop |> clean_names() |>
  filter(country_name=="Denmark") |> 
  select(year, population) |> drop_na()
```

```{r}
#On the observed scale
denmark_ts<-ggplot(den_data, aes(x=year, y=population))+
  geom_point()+
  xlab("Year")+
  ylab("Human count (N)")+
  xlim(1949, 2023)+
  ggtitle("Human Population in Denmark (1950-2022)")+
  theme_bw()

denmark_ts
```

```{r}
#on the transformed log scale
denmark_ts_log<-ggplot(den_data, aes(x=year, y=population))+
  geom_point()+
  xlab("Year")+
  ylab("Human count (N)")+
  scale_y_continuous(trans="log")+ # log scale for linear transformation
  xlim(1949, 2023)+
  ggtitle("Human Population in Denmark (1950-2022)")+
  theme_bw()

denmark_ts_log
```

```{r}
denmark.lm_fit<- lm(log(population)~year,data=den_data)
summary(denmark.lm_fit) #how well does it fit? 

NO<-exp(denmark.lm_fit$coefficients[1])
r<-denmark.lm_fit$coefficients[2]
```

```{r}
# On the transformed scale
denmark_ts_model_trans<-ggplot(den_data, aes(x=year, y=population))+
  geom_point()+
  xlab("Year")+
  ylab("Human count (N)")+
  scale_y_continuous(trans="log")+
  geom_smooth(method="lm",color="blue")+
  xlim(1949, 2023)+
  ggtitle("Human Population in Denmark (1950-2022)")+
  theme_bw()

denmark_ts_model_trans
```

```{r}
# On the observed scale

# Predicted population values from the model
den_predicted_df <- data.frame(den_pop_pred = predict(denmark.lm_fit, den_data), year=den_data$year)
den_predicted_df<- den_predicted_df |> mutate(N_est=exp(den_pop_pred)) #transformed back to observed popualtion estimates


denmark_ts_model<-ggplot(den_data, aes(x=year, y=population))+
  geom_point(color="black")+
  geom_line(color='red',data = den_predicted_df, aes(x=year, y=N_est))+
  xlab("Year")+
  ylab("Human count (N)")+
  xlim(1949, 2023)+
  ggtitle("Human Population in Denmark (1950-2022)")+
  theme_bw()

denmark_ts_model
```


**Logistic Growth of Human and Bison Populations Example**

```{r}
#transform year to time for world pop data
w_pop_data <- w_pop_data |> mutate(time = year - year[1])
```

```{r}
#using the growthcurver package
#here: https://cran.r-project.org/web/packages/growthcurver/vignettes/Growthcurver-vignette.html

human_fit<-SummarizeGrowth(w_pop_data$time,w_pop_data$population)
human_fit
plot(human_fit) #what is funny with this graph?  N at 0 is what?  
w_pop_data$population[1] # need to add this to the y population values.  
est_K_human<-human_fit$vals$k + w_pop_data$population[1] #estimated K for the data
```

```{r}
#bison data
bison_data<-read_csv(here("Portfolio1","bison.csv"))

#transform year to time
bison_data <- bison_data |> mutate(time = year - year[1])
```

```{r}
bison_fit<-SummarizeGrowth(bison_data$time,bison_data$bison)
bison_fit
plot(bison_fit)
est_K_bison<-bison_fit$vals$k + bison_data$bison[1] #estimated K for the data
```

**Logistic Growth of Denmark Population**

```{r}
#transform year to time
den_data <- den_data |> mutate(time = year - year[1])
```

```{r}
denmark_fit<-SummarizeGrowth(den_data$time,den_data$population)
denmark_fit
plot(denmark_fit) #what is funny with this graph?  N at 0 is what?  
den_data$population[1] # need to add this to the y population values.  
est_K_denmark<-denmark_fit$vals$k + den_data$population[1] #estimated K for the data
```





